---
title: "RDS_curations"
format: html
editor: visual
---

```{r}
library(phyloseq)
library(speedyseq)
library(Biostrings)
library(tidyverse)
library(microbiomeutilities)
library(stringr)
```


## Metadata

```{r}
metadata <- read_csv("/Users/fst584/Documents/MATRIX/core_microbiome/ASV_processing/Metadata/metadata_all.csv")
```

### Add rerun samples to metadata

```{r}
rerun <- read_csv("/Users/fst584/Desktop/new_run/BRLUV_DR8PG_rerun_samples.csv")

metadata_dummy <- metadata %>% 
  mutate(run_name = str_extract(run_ID, "(?<=_)[^_]*$")) %>% 
  select(!c(forward_fq, reverse_fq, run_ID))

rerun <- rerun %>% 
  mutate(run_name = str_extract(run_ID, "(?<=_)[^_]*$")) %>% 
  left_join(metadata_dummy, by = c("seq_sample_ID", "run_name")) %>% 
  select(!c(run_name))


metadata <- bind_rows(metadata,rerun)
```

### Create rownames and purpose column
#### Make seq run samples in DWDQH and JRV78 unique

```{r}
# Add 'rowname' column (the unique identifier in count table)
metadata <- metadata %>% 
  filter(!is.na(forward_fq)) %>% 
  mutate(rowname = paste0(run_ID, "/", forward_fq)) 

# Find duplicate rows (i.e samples that are counted twice, both as sample and as seq_control)
dupes <- metadata %>% count(rowname) %>% filter(n>1)

# Create purpose columns and remove duplicate samples
metadata <- metadata%>% 
  mutate(run_name = str_extract(run_ID, "(?<=_)[^_]*$")) %>% 
  mutate(IS_SEQ_CONTROL = ifelse(str_detect(suggested_correct_seq_sample_ID ,pattern=run_name), "Y",
                                 ifelse(rowname %in% dupes$rowname, "Y", "N"))) %>% 
  mutate(IS_SAMPLE = ifelse((is.na(LY) | str_detect(LY, pattern = "0000")), "N",
                            ifelse(str_detect(suggested_correct_seq_sample_ID ,pattern=run_name), "N", "Y"))) %>% 
  mutate(IS_LIB_CONTROL = ifelse(str_detect(LY, pattern = "0000"), "Y", "N")) %>% 
  filter(!((rowname %in% dupes$rowname) & IS_SAMPLE == "N")) %>% 
  mutate(sample = rowname)
```

### Recreate metadata columns

```{r}
metadata <- metadata %>% 
  mutate(TYPE = ifelse(str_detect(source, pattern = "crushed"), "crushed",
                       ifelse(str_detect(source, pattern = "shotgun_amplicon"), "shotgun_amplicon",
                              "washed"))) %>% 
  mutate(location = SP_location) %>% 
  mutate(spray_fungicide = coalesce(spray, spray_fungicide),
         spray_fungicide = case_when(
           spray_fungicide%in% c("y", "yes") ~ "yes",
           spray_fungicide %in% c("n", "no")  ~ "no",
           TRUE ~ NA_character_)) %>% 
  mutate(fertiliser = coalesce(SP_fertiliser, SP_fertilizer)) %>% 
  select(!c("cultivar", "Cultivar")) %>% 
  rename(cultivar = SP_cultivar) %>% 
  mutate(Date = case_when(
    source == "2022_washed_amplicon" ~ as.Date(sample_date, format = "%d/%m/%y"),
    source %in% c("2020_washed_amplicon",
                  "2021_crushed_amplicon",
                  "2022_crushed_amplicon",
                  "control_crushed_amplicon",
                  "2022_washed_shotgun_amplicon",
                  "2023_crushed_amplicon_16S",
                  "2023_crushed_amplicon_ITS",
                  "2023_washed_shotgun_amplicon") ~ as.Date(sample_date, format = "%Y-%m-%d %H:%M:%S")
  )) %>% 
  mutate(sample_day_of_the_year = Date-as.Date(paste0("1/1/", YEAR), format = "%d/%m/%Y"))
```



### Create sample_data object

```{r}
metadata_sample_data <- metadata %>% column_to_rownames() %>% sample_data()
```



## ITS

### Load ITS phyloseq

```{r}
ps_ITS <- readRDS("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/ITS/combined_physeq.rds")

sample_data(ps_ITS) <- metadata_sample_data
```

### Save DNA as biostrings and name ASVs ASV1, ASV2 etc.

```{r}
dna_ITS <- Biostrings::DNAStringSet(taxa_names(ps_ITS))
names(dna_ITS) <- taxa_names(ps_ITS)
ps_ITS <- merge_phyloseq(ps_ITS, dna_ITS)
taxa_names(ps_ITS) <- paste0("ASV", seq(ntaxa(ps_ITS)))
```


### Get LULU results

```{r}
LULU_ITS <- readRDS("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/ITS/lulu_curated.rds")

otu_table(ps_ITS) <- t(LULU_ITS$curated_table) %>% otu_table(taxa_are_rows = F)
```

### Add taxonomy to ASV names

```{r}
new_ITS_taxtable <- ps_ITS %>%
  tax_table %>%
  as.data.frame() %>%
  mutate(Species = ifelse(is.na(Genus),
                          Species,
                          ifelse(is.na(Species),
                                 Genus,
                                 paste0(gsub("g__", "", Genus), "_", gsub("s__", "", Species)
                                        ))))


tax_table(ps_ITS) <- as.matrix(new_ITS_taxtable)

ps_ITS <- microbiomeutilities::format_to_besthit(ps_ITS)

ps_ITS <- ps_ITS %>% subset_taxa(!(Phylum  %in% c("k__Fungi", "k__Unclassified")))
```


### Get reports

```{r}
track_ITS <- read_tsv("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/ITS/combined_read_track.tsv")

track_ITS <- track_ITS %>% 
  mutate(lost_trimming = raw - trimmed,
         lost_denoising = trimmed - denoisedF,
         lost_merging = denoisedF - merged,
         lost_chimera = merged - nonchim,
         lost_total = raw - nonchim) %>% 
  mutate(lost_trimming_pct = lost_trimming/raw,
         lost_denoising_pct = lost_denoising/trimmed,
         lost_merging_pct = lost_merging/denoisedF,
         lost_chimera_pct = lost_chimera/merged,
         lost_total_pct = lost_total/raw) %>% 
  mutate(sample = paste0(seq_run, "/", sample, "_R1_001.fastq.gz"))

ps_ITS <- ps_ITS %>% speedyseq::left_join_sample_data(track_ITS)

# count total reads in final object
sums_ITS <- sample_sums(ps_ITS) %>% data.frame
colnames(sums_ITS) <- c("ASV_reads")
sums_ITS$sample <- rownames(sums_ITS)

ps_ITS <- ps_ITS %>% speedyseq::left_join_sample_data(sums_ITS)

ps_ITS <- ps_ITS %>% mutate_sample_data(pass_filter = ifelse((lost_total_pct>=90 | ASV_reads<=1000), "no", "yes"))
```


### Calculate alpha diversity metrics

```{r}
ps_ITS_counts <- ps_ITS %>% otu_table() %>% data.frame()

diversity_ITS <- ps_ITS_counts %>% 
  rowwise() %>%
  mutate(berger_parker = max(c_across(everything())) / sum(c_across(everything())),
         observed = sum(c_across(!berger_parker) > 0)) %>% 
  select(berger_parker, observed)

diversity_ITS$sample <- rownames(ps_ITS_counts)

ps_ITS <- ps_ITS %>% speedyseq::left_join_sample_data(diversity_ITS)
```


### Save curated RDS

```{r}
saveRDS(ps_ITS, "/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/ITS/ITS_ASVs_curated.rds")
```






## 16S

### Load 16S phyloseq

```{r}
ps_16S <- readRDS("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/16S/combined_physeq.rds")

sample_data(ps_16S) <- metadata_sample_data
```

### Save DNA as biostrings and name ASVs ASV1, ASV2 etc.

```{r}
dna_16S <- Biostrings::DNAStringSet(taxa_names(ps_16S))
names(dna_16S) <- taxa_names(ps_16S)
ps_16S <- merge_phyloseq(ps_16S, dna_16S)
taxa_names(ps_16S) <- paste0("ASV", seq(ntaxa(ps_16S)))
```


### Get LULU results

```{r}
LULU_16S <- readRDS("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/16S/lulu_curated.rds")

otu_table(ps_16S) <- t(LULU_16S$curated_table) %>% otu_table(taxa_are_rows = F)
```

### Add taxonomy to ASV names

```{r}
ps_16S <- microbiomeutilities::format_to_besthit(ps_16S)

ps_16S <- ps_16S %>% subset_taxa(!(Phylum  %in% c("k__Bacteria", "k__Unclassified")))
ps_16S <- phyloseq::subset_taxa(ps_16S, (Family!= "Mitochondria" | is.na(Family)) & (Order!="Chloroplast" | is.na(Order)))
```


### Get reports

```{r}
track_16S <- read_tsv("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/16S/combined_read_track.tsv")

track_16S <- track_16S %>% 
  mutate(lost_trimming = raw - trimmed,
         lost_denoising = trimmed - denoisedF,
         lost_merging = denoisedF - merged,
         lost_chimera = merged - nonchim,
         lost_total = raw - nonchim) %>% 
  mutate(lost_trimming_pct = lost_trimming/raw,
         lost_denoising_pct = lost_denoising/trimmed,
         lost_merging_pct = lost_merging/denoisedF,
         lost_chimera_pct = lost_chimera/merged,
         lost_total_pct = lost_total/raw) %>% 
  mutate(sample = paste0(seq_run, "/", sample, "_R1_001.fastq.gz"))

ps_16S <- ps_16S %>% speedyseq::left_join_sample_data(track_16S)

sums_16S <- sample_sums(ps_16S) %>% data.frame
colnames(sums_16S) <- c("ASV_reads")
sums_16S$sample <- rownames(sums_16S)

ps_16S <- ps_16S %>% speedyseq::left_join_sample_data(sums_16S)

ps_16S <- ps_16S %>% mutate_sample_data(pass_filter = ifelse((lost_total_pct>=90 | ASV_reads<=1000), "no", "yes"))
```

### Calculate alpha diversity metrics

```{r}
ps_16S_counts <- ps_16S %>% otu_table() %>% data.frame()

diversity_16S <- ps_16S_counts %>% 
  rowwise() %>%
  mutate(berger_parker = max(c_across(everything())) / sum(c_across(everything())),
         observed = sum(c_across(!berger_parker) > 0)) %>% 
  select(berger_parker, observed)

diversity_16S$sample <- rownames(ps_16S_counts)

ps_16S <- ps_16S %>% speedyseq::left_join_sample_data(diversity_16S)
```


### Save curated RDS

```{r}
saveRDS(ps_16S, "/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/16S/16S_ASVs_curated.rds")
```









## Space for investigating metadata

```{r}
ps_ITS <- readRDS("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/ITS/ITS_ASVs_curated.rds")
ps_16S <- readRDS("/Users/fst584/Documents/MATRIX/core_microbiome/processed_data/16S/16S_ASVs_curated.rds")

data_ITS <- ps_ITS %>% sample_data %>% data.frame
data_16S <- ps_16S %>% sample_data %>% data.frame
```



```{r}
ps_16S %>% subset_samples(YEAR == "2021") %>% sample_data %>% data.frame
ps_16S %>% sample_names %>% data.frame %>% filter(str_detect(., pattern = "21-"))
```



```{r}
metadata %>% filter(!str_detect(forward_fq, "ITS|16S")) %>% filter(amplicon == "ITS")

metadata %>% filter(!str_detect(forward_fq, "ITS|16S")) %>% filter(amplicon == "16S")
```


